---
title: "Why Have House Prices in Turkey Gone Up in the Last Ten Years?"
author:
  - name: "Ömer Furkan Demirci"
  - name: "Fatih Alper Vural"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    fig-width: 9
    fig-height: 5
    fig-cap-location: bottom
    df-print: paged
  pdf:
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    margin-left: 2cm
    margin-right: 2cm
    margin-top: 2cm
    margin-bottom: 2cm
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

------------------------------------------------------------------------

## Abstract {.unnumbered}

Housing affordability has emerged as one of the most pressing socioeconomic challenges in Türkiye over the past decade. This study investigates the key drivers of the sharp and sustained rise in housing prices between 2014 and 2024, drawing on four official monthly time-series datasets obtained from the Central Bank of the Republic of Türkiye (TCMB/EVDS): the House Price Index (HPI), housing sales statistics, USD/EUR exchange rates, and credit card sectoral expenditure data. Through exploratory analysis, correlation testing, regional comparison, and multiple linear regression modelling, we find that exchange rate depreciation and broad consumer spending inflation are the dominant macroeconomic forces behind rising house prices, while sales volumes display an inverse relationship with price levels — consistent with demand compression at high price levels. Regional heterogeneity is substantial: İstanbul and western metropolitan areas lead price growth, whereas interior Anatolian regions lag considerably. Our model explains approximately 95% of the variance in the national HPI, suggesting that macroeconomic indicators are strong predictors of housing price trajectories in Türkiye.

------------------------------------------------------------------------

# Project Overview and Scope

We want to look into the following question for this project:\
\
**Why have house prices in Turkey gone up in the last ten years?**\
\
To answer this question, we made a dataset using four monthly time-series datasets from the Central Bank of the Republic of Türkiye's (TCMB) Electronic Data Delivery System (EVDS). These datasets have:\

-   Housing Price Index

-   Sales of Houses

-   Exchange Rates (USD/TRY and EUR/TRY)

-   Credit Card Sectoral Expenditures\
    \

We chose this topic because the cost of housing has become one of the most important social and economic issues in Türkiye. It is important to know what caused this rise in order to make good decisions about policy and the economy. This problem is also a good fit for a data analytics approach because it needs to combine multiple datasets, do preprocessing, and look at how things change over time and in different areas.

We want to find the main things that cause housing prices to go up. Specifically, we want to look at:

-   How exchange rates affect the price of homes

-   The link between changes in housing prices and demand (sales)

-   The function of consumption behavior (credit card expenditure) as an indicator of demand conditions.

-   Differences in how housing works in different regions

# Data

You can see Our data in R Data format below.

[Project Data](https://github.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/blob/e273d93492dc0a2ea56f85a472a398fe4909c2d7/Data/Project_Data.RData)

```{r}
url <- "https://raw.githubusercontent.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/main/Data/Project_Data.RData"
download.file(url, destfile = "Project_Data.RData", mode = "wb")
load("Project_Data.RData")

ls()
```

## Data Source

The data used in this project were obtained from the **Electronic Data Delivery System (EVDS)** of the Central Bank of the Republic of Türkiye (TCMB). EVDS provides official, high-frequency economic data and is widely used for academic and professional analysis.

The project is based on four datasets:

-   **House Price Index (HPI)**\
    Monthly housing price index at national and regional levels

-   **House Sales Statistics**\
    Monthly housing sales data at the provincial level

-   **Exchange Rates (USD, EUR)**\
    Monthly exchange rate data

-   **Credit Card Sectoral Expenditures**\
    Monthly sector-based credit card spending data

## General Information About Data

The processed dataset consists of three main analytical tables:

### Exchange Rate Table

-   **Description**:\
    This table contains the columns Year, Month, USD Conversion Rate, and EUR Conversion Rate. Represents monthly exchange rate movements. These variables are used as indicators of macroeconomic conditions and cost-side pressures that may affect housing prices.

-   In general, the exchange rate variables show a strong upward trend over time rather than a stable distribution.

    ```{r}
    library(dplyr)
    library(ggplot2)
    usd_eur <- Conversion_Rates %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(usd_eur, aes(x = Date)) +
      geom_line(aes(y = USD, color = "USD")) +
      geom_line(aes(y = EUR, color = "EUR")) +
      labs(
        title = "Exchange Rates Over Time",
        x = "Date",
        y = "Value",
        color = "Series"
      ) +
      theme_minimal()
    ```

### Credit Card Expenditure Table

-   **Description**:

-   This table contains Year, Month, Expense Type, and Quantity.\

-   Expense Type identifies the spending category, and Quantity gives the corresponding spending amount for that month.

-   This table contains sectoral spending amounts derived from credit card transactions.

-   The values in this table are expected to differ substantially across categories. Some categories such as retail, food, fuel, and e-commerce are likely to have much larger values than others.

    ```{r}
    library(dplyr)
    library(ggplot2)
    expense_quantity <- Expense_Quantity %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      expense_quantity %>%
        filter(Expense_Type %in% c("E-Alışveriş", "Çeşitli Gıda", "Giyim ve Aksesuar", "Benzin ve Yakıt İstasyonları")),
      aes(x = Date, y = Quantity, color = Expense_Type)
    ) +
      geom_line() +
      labs(
        title = "Credit Card Expenditures for Selected Categories",
        x = "Date",
        y = "Quantity",
        color = "Expense Type"
      ) +
      theme_minimal()
    ```

### Housing Panel Dataset

-   **Description**:\
    This is the main table for the project and contains Year, Month, Region, House Price Index, and Total Sales.

    Region represents the housing price index regions as in NUTS2 level, House Price Index shows the monthly house price index for that region, and Total Sales gives the aggregated monthly house sales value for the same region.

    ```{r}
    library(dplyr)
    library(ggplot2)

    konut_index_sales <- Konut_Index_Sales %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      konut_index_sales %>% filter(Region == "Türkiye"),
      aes(x = Date, y = Konut_Endeksi)
    ) +
      geom_line() +
      labs(
        title = "House Price Index in Türkiye Over Time",
        x = "Date",
        y = "House Price Index"
      ) +
      theme_minimal()
    ```

```{r}
library(dplyr)
library(ggplot2)

ggplot(
  Konut_Index_Sales %>%
    mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month))) %>%
    filter(Region %in% c("Türkiye", "İstanbul", "İzmir", "Ankara", "Adana, Mersin")),
  aes(x = Date, y = Total_Sales, color = Region)
) +
  geom_line() +
  labs(
    title = "Total House Sales for Selected Regions",
    x = "Date",
    y = "Total Sales",
    color = "Region"
  ) +
  theme_minimal()
```

## Potential Analyses

-   time-series plots of house prices, exchange rates, and sales,
-   regional comparisons of house price growth,
-   correlations between exchange rates and the house price index,
-   relationships between sales volume and housing prices,
-   category-based spending trends over time,

## Reason of Choice

Housing prices are one of the most important economic and social issues in Türkiye. The rapid increase in housing prices has significant implications for affordability, investment decisions, and economic stability.

All datasets are:

-   Consistent in frequency (monthly)

-   Long enough for trend analysis

-   Rich enough for regional comparison

    This makes it easy for analyzing.

## Preprocessing

The raw EVDS datasets were provided in Excel format and contained different structures (wide vs. long) and aggregation levels (province vs. region). We performed preprocessing in Python using pandas to standardize date formats, reshape wide datasets into long format, map categorical codes to meaningful labels, and align all datasets at a consistent monthly and regional level. In particular, house sales data were converted from province-level columns to regional aggregates compatible with the House Price Index. The final result consists of three clean analytical tables ready for further analysis.

Another key issue was that EVDS datasets use **coded column names** (e.g., `KT2`, `T58`, `TR62`) instead of descriptive labels. To make the data interpretable and consistent across sources, we created mapping dictionaries in Python to translate these codes into meaningful categories, provinces, and regions.

Here you can find our python code.\
\

``` py
# We explained the data and the output tables to AI. It gave us this code's first version. Then we fixed it according to our needs
import pandas as pd
from pathlib import Path

downloads = Path.home() / "Downloads"

# Files
kurlar_file = downloads / "Kurlar.xlsx"
kredi_file = downloads / "Kredi Harcama.xlsx"
konut_satis_file = downloads / "Konut Satış.xlsx"
konut_fiyat_file = downloads / "Konut Fiyat Endexi.xlsx"

output_file = downloads / "evds_preprocessed_tables_fixed.xlsx"

# -----------------------------
# Helper
# -----------------------------
def split_year_month(df):
    df["Tarih"] = pd.to_datetime(df["Tarih"], format="%Y-%m")
    df["Year"] = df["Tarih"].dt.year
    df["Month"] = df["Tarih"].dt.month
    return df

# -----------------------------
# Read
# -----------------------------
kurlar = split_year_month(pd.read_excel(kurlar_file))
kredi = split_year_month(pd.read_excel(kredi_file))
satis = split_year_month(pd.read_excel(konut_satis_file))
kfe = split_year_month(pd.read_excel(konut_fiyat_file))

# -----------------------------
# TABLE 1: Exchange Rates
# -----------------------------
table_kurlar = (
    kurlar[["Year", "Month", "TP_DK_USD_S_YTL", "TP_DK_EUR_S_YTL"]]
    .rename(columns={
        "TP_DK_USD_S_YTL": "USD",
        "TP_DK_EUR_S_YTL": "EUR"
    })
)

# -----------------------------
# TABLE 2: Credit Expenditures
# -----------------------------
expense_type_map = {
    "TP_KKHARTUT_KT2": "Car Rental",
    "TP_KKHARTUT_KT3": "Car Sales/Service",
    "TP_KKHARTUT_KT4": "Fuel",
    "TP_KKHARTUT_KT5": "Food",
    "TP_KKHARTUT_KT6": "Direct Marketing",
    "TP_KKHARTUT_KT7": "Education",
    "TP_KKHARTUT_KT8": "Electronics",
    "TP_KKHARTUT_KT9": "Clothing",
    "TP_KKHARTUT_KT10": "Airlines",
    "TP_KKHARTUT_KT11": "Services",
    "TP_KKHARTUT_KT12": "Accommodation",
    "TP_KKHARTUT_KT13": "Social",
    "TP_KKHARTUT_KT14": "Entertainment",
    "TP_KKHARTUT_KT15": "Jewelry",
    "TP_KKHARTUT_KT16": "Retail",
    "TP_KKHARTUT_KT17": "Furniture",
    "TP_KKHARTUT_KT18": "Construction",
    "TP_KKHARTUT_KT19": "Health",
    "TP_KKHARTUT_KT20": "Travel",
    "TP_KKHARTUT_KT21": "Insurance",
    "TP_KKHARTUT_KT22": "Telecom",
    "TP_KKHARTUT_KT23": "Hardware",
    "TP_KKHARTUT_KT24": "Food Service",
    "TP_KKHARTUT_KT25": "Misc",
    "TP_KKHARTUT_KT26": "Other",
    "TP_KKHARTUT_KT50": "E-Commerce"
}

kredi_cols = [c for c in expense_type_map if c in kredi.columns]

table_kredi = (
    kredi.melt(
        id_vars=["Year", "Month"],
        value_vars=kredi_cols,
        var_name="Code",
        value_name="Quantity"
    )
    .assign(Expense_Type=lambda x: x["Code"].map(expense_type_map))
    [["Year", "Month", "Expense_Type", "Quantity"]]
)

# -----------------------------
# TABLE 3: Housing Panel
# -----------------------------
provinces = [
    "Adana","Adıyaman","Afyonkarahisar","Ağrı","Aksaray","Amasya","Ankara",
    "Antalya","Ardahan","Artvin","Aydın","Balıkesir","Bartın","Batman",
    "Bayburt","Bilecik","Bingöl","Bitlis","Bolu","Burdur","Bursa","Çanakkale",
    "Çankırı","Çorum","Denizli","Diyarbakır","Düzce","Edirne","Elazığ",
    "Erzincan","Erzurum","Eskişehir","Gaziantep","Giresun","Gümüşhane",
    "Hakkâri","Hatay","Iğdır","Isparta","İstanbul","İzmir","Kahramanmaraş",
    "Karabük","Karaman","Kars","Kastamonu","Kayseri","Kırıkkale","Kırklareli",
    "Kırşehir","Kilis","Kocaeli","Konya","Kütahya","Malatya","Manisa",
    "Mardin","Mersin","Muğla","Muş","Nevşehir","Niğde","Ordu","Osmaniye",
    "Rize","Sakarya","Samsun","Siirt","Sinop","Sivas","Şanlıurfa","Şırnak",
    "Tekirdağ","Tokat","Trabzon","Tunceli","Uşak","Van","Yalova","Yozgat",
    "Zonguldak"
]

code_map = {"TP_AKONUTSAT1_K1": provinces[0]}
for i in range(2, 82):
    code_map[f"TP_AKONUTSAT1_T{i}"] = provinces[i-1]

sales_cols = [c for c in satis.columns if c in code_map]

sales_long = (
    satis.melt(
        id_vars=["Year", "Month"],
        value_vars=sales_cols,
        var_name="Code",
        value_name="Sales"
    )
    .assign(Province=lambda x: x["Code"].map(code_map))
)

# Region mapping (example simplified)
region_map = {
    "Adana":"TR62","Mersin":"TR62","Ankara":"TR51","İstanbul":"TR10"
}

sales_long["Region_Code"] = sales_long["Province"].map(region_map)

sales_region = (
    sales_long.groupby(["Year","Month","Region_Code"])["Sales"].sum().reset_index()
)

kfe_map = {
    "TP_KFE_TR62":"TR62",
    "TP_KFE_TR51":"TR51",
    "TP_KFE_TR10":"TR10"
}

kfe_cols = [c for c in kfe_map if c in kfe.columns]

kfe_long = (
    kfe.melt(
        id_vars=["Year","Month"],
        value_vars=kfe_cols,
        var_name="Code",
        value_name="House_Price_Index"
    )
    .assign(Region_Code=lambda x: x["Code"].map(kfe_map))
)

table_konut = kfe_long.merge(
    sales_region,
    on=["Year","Month","Region_Code"]
)

# -----------------------------
# SAVE
# -----------------------------
with pd.ExcelWriter(output_file) as writer:
    table_kurlar.to_excel(writer, sheet_name="USD_EUR", index=False)
    table_kredi.to_excel(writer, sheet_name="Expense", index=False)
    table_konut.to_excel(writer, sheet_name="Housing", index=False)
```

# Analysis

This section presents the full analytical workflow: exploratory data analysis, trend decomposition, correlation analysis, regional comparison, and predictive modelling. All code is collapsible for readability; outputs and narrative explanations accompany each step.

## Exploratory Data Analysis

We begin by loading the data and performing a structured overview of each of the three tables.

```{r}
#| label: load-data

library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)

# Load data
load("Project_Data.RData")

# Build date columns
cr <- Conversion_Rates %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

eq <- Expense_Quantity %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

ks <- Konut_Index_Sales %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))
```

### Summary Statistics

#### Exchange Rates

```{r}
#| label: eda-exchange-summary

cr_summary <- cr %>%
  select(USD, EUR) %>%
  summarise(
    across(everything(), list(
      Min    = min,
      Mean   = mean,
      Median = median,
      Max    = max,
      SD     = sd
    ), .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(), names_to = c("Currency", "Stat"), names_sep = "_") %>%
  pivot_wider(names_from = Stat, values_from = value)

kable(cr_summary, digits = 2, caption = "Summary Statistics – Exchange Rates (TRY)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The USD/TRY rate grew from a minimum near 2 TRY to over 30 TRY across the sample period, illustrating the scale of currency depreciation. The EUR/TRY series followed a similar trajectory. The high standard deviations relative to means confirm that these series are strongly trended rather than stationary.

#### House Price Index and Sales (National)

```{r}
#| label: eda-hpi-summary

ks_turkey <- ks %>% filter(Region == "Türkiye")

hpi_summary <- ks_turkey %>%
  summarise(
    N          = n(),
    HPI_Min    = min(Konut_Endeksi, na.rm = TRUE),
    HPI_Mean   = mean(Konut_Endeksi, na.rm = TRUE),
    HPI_Median = median(Konut_Endeksi, na.rm = TRUE),
    HPI_Max    = max(Konut_Endeksi, na.rm = TRUE),
    HPI_SD     = sd(Konut_Endeksi, na.rm = TRUE),
    Sales_Min  = min(Total_Sales, na.rm = TRUE),
    Sales_Mean = mean(Total_Sales, na.rm = TRUE),
    Sales_Max  = max(Total_Sales, na.rm = TRUE),
    Sales_SD   = sd(Total_Sales, na.rm = TRUE)
  )

kable(t(hpi_summary), col.names = "Value", caption = "Summary – National HPI and Sales") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The national HPI has grown dramatically, with the maximum value being many multiples of the minimum. Sales volumes display high variability, reflecting seasonal patterns and market shifts.

#### Credit Card Expenditure (Top Categories)

```{r}
#| label: eda-credit-summary

top_cats <- eq %>%
  group_by(Expense_Type) %>%
  summarise(Total = sum(Quantity, na.rm = TRUE)) %>%
  arrange(desc(Total)) %>%
  slice_head(n = 8)

kable(top_cats, digits = 0,
      col.names = c("Expense Category", "Total Spending (Million TRY)"),
      caption = "Top 8 Credit Card Expenditure Categories (Cumulative)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Retail trade, food, and e-commerce consistently rank as the largest credit card spending categories. Furniture and construction-related categories, while smaller in absolute terms, are of particular interest given their direct linkage to the housing market.

### Distribution Analysis

```{r}
#| label: eda-distributions
#| fig-cap: "Distribution of key variables"
#| fig-height: 4

p1 <- ggplot(ks_turkey, aes(x = Konut_Endeksi)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  labs(title = "HPI Distribution", x = "House Price Index", y = "Count") +
  theme_minimal()

p2 <- ggplot(cr, aes(x = USD)) +
  geom_histogram(bins = 30, fill = "#d7191c", color = "white", alpha = 0.85) +
  labs(title = "USD/TRY Distribution", x = "USD Rate", y = "Count") +
  theme_minimal()

library(patchwork)
p1 + p2
```

Both the HPI and the exchange rate show right-skewed, strongly non-normal distributions — a natural consequence of exponential growth trajectories. This motivates the use of log-transformations in the regression modelling step.

## Trend Analysis

### House Price Index – National Trend

```{r}
#| label: trend-hpi
#| fig-cap: "National House Price Index (2014–2024) with trend line"

ggplot(ks_turkey, aes(x = Date, y = Konut_Endeksi)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  geom_smooth(method = "loess", se = TRUE, color = "#d7191c",
              fill = "#fdae61", alpha = 0.25, linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "National House Price Index Over Time",
    subtitle = "Monthly data from TCMB-EVDS; LOESS trend overlay",
    x = "Date", y = "House Price Index"
  ) +
  theme_minimal(base_size = 11)
```

The national HPI shows a gradual increase through 2019, followed by an accelerating surge from 2020 onward. Three distinct phases are visible: a moderate growth phase (2014–2019), an early acceleration phase (2020–2021), and a sharp exponential surge (2022–2024), closely coinciding with episodes of severe TRY depreciation.

### Exchange Rate Trends

```{r}
#| label: trend-fx
#| fig-cap: "USD/TRY and EUR/TRY exchange rates over time"

cr_long <- cr %>%
  pivot_longer(cols = c(USD, EUR), names_to = "Currency", values_to = "Rate")

ggplot(cr_long, aes(x = Date, y = Rate, color = Currency)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("USD" = "#2c7bb6", "EUR" = "#1a9641")) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Exchange Rate Movements (USD/TRY, EUR/TRY)",
    subtitle = "Data source: TCMB-EVDS",
    x = "Date", y = "TRY per Foreign Currency"
  ) +
  theme_minimal(base_size = 11)
```

Both rates track each other closely and display the same phase structure as the HPI, providing initial visual evidence of co-movement. The sharpest depreciations occurred in 2021–2022 and again in 2023, corresponding directly to the most aggressive phases of HPI growth.

### Housing Sales Trend

```{r}
#| label: trend-sales
#| fig-cap: "National monthly housing sales with 12-month moving average"

ks_turkey_ma <- ks_turkey %>%
  arrange(Date) %>%
  mutate(MA12 = zoo::rollmean(Total_Sales, k = 12, fill = NA, align = "right"))

ggplot(ks_turkey_ma, aes(x = Date)) +
  geom_col(aes(y = Total_Sales), fill = "#abd9e9", alpha = 0.6) +
  geom_line(aes(y = MA12), color = "#d7191c", linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Monthly Housing Sales – Türkiye",
    subtitle = "Bars = monthly volume; Red line = 12-month moving average",
    x = "Date", y = "Number of Units Sold"
  ) +
  theme_minimal(base_size = 11)
```

Sales volumes show pronounced seasonality (peaks in spring and year-end) and a secular downward trend from 2022 onward, which is paradoxical relative to rising prices. This divergence is consistent with supply-demand dynamics in which high prices price out buyers, reducing transactional volume while asset prices continue to inflate — a phenomenon often observed in high-inflation real estate markets.

### Credit Card Spending Trends

```{r}
#| label: trend-credit
#| fig-cap: "Annual total credit card expenditure by selected categories"

annual_exp <- eq %>%
  group_by(Year, Expense_Type) %>%
  summarise(Total = sum(Quantity, na.rm = TRUE), .groups = "drop")

top5_cats <- annual_exp %>%
  group_by(Expense_Type) %>%
  summarise(GrandTotal = sum(Total)) %>%
  arrange(desc(GrandTotal)) %>%
  slice_head(n = 5) %>%
  pull(Expense_Type)

ggplot(annual_exp %>% filter(Expense_Type %in% top5_cats),
       aes(x = Year, y = Total / 1e6, color = Expense_Type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Annual Credit Card Expenditure – Top 5 Categories",
    subtitle = "Values in billion TRY",
    x = "Year", y = "Expenditure (Billion TRY)", color = "Category"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

All spending categories show exponential nominal growth from 2020 onward. Importantly, this growth is almost certainly nominal rather than real — it reflects the broad inflationary environment rather than genuine increases in transaction volumes — which itself reinforces the causal chain from inflation and currency depreciation to rising housing prices.

## Regional Analysis

### HPI Growth by Region

```{r}
#| label: regional-hpi
#| fig-cap: "House Price Index trajectories for selected NUTS2 regions"

selected_regions <- c("Türkiye", "İstanbul", "Ankara", "İzmir",
                      "Adana, Mersin", "Konya, Karaman",
                      "Antalya, Burdur, Isparta")

ggplot(
  ks %>% filter(Region %in% selected_regions),
  aes(x = Date, y = Konut_Endeksi, color = Region)
) +
  geom_line(linewidth = 0.75, alpha = 0.9) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "House Price Index by Region",
    subtitle = "Selected NUTS2 regions, TCMB-EVDS",
    x = "Date", y = "House Price Index", color = "Region"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

```{r}
#| label: regional-growth-table

# Calculate total index growth per region (first to last observation)
region_growth <- ks %>%
  group_by(Region) %>%
  arrange(Date) %>%
  summarise(
    First_HPI = first(Konut_Endeksi),
    Last_HPI  = last(Konut_Endeksi),
    Growth_pct = (Last_HPI / First_HPI - 1) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(Growth_pct)) %>%
  filter(Region != "Türkiye")

kable(region_growth,
      digits = 1,
      col.names = c("Region", "First HPI", "Latest HPI", "Growth (%)"),
      caption = "House Price Index Growth by NUTS2 Region (Full Sample)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

İstanbul and western coastal regions (Antalya, İzmir) consistently lead in absolute HPI levels. However, the growth rate ranking is more nuanced — some central Anatolian regions recorded comparable or even higher percentage growth from a lower base. This regional heterogeneity suggests that both demand-pull (migration, tourism) and supply-side constraints vary considerably across geography.

### Heatmap of Regional Sales

```{r}
#| label: regional-sales-heatmap
#| fig-cap: "Annual housing sales by region (heat map)"
#| fig-height: 7

annual_sales_region <- ks %>%
  filter(Region != "Türkiye") %>%
  group_by(Year, Region) %>%
  summarise(Annual_Sales = sum(Total_Sales, na.rm = TRUE), .groups = "drop")

ggplot(annual_sales_region, aes(x = Year, y = Region, fill = Annual_Sales)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#ffffb2", high = "#bd0026", labels = comma) +
  labs(
    title = "Annual Housing Sales by Region",
    subtitle = "Darker = higher sales volume",
    x = "Year", y = NULL, fill = "Units Sold"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 7.5))
```

İstanbul dominates sales volume by a wide margin across all years, followed by Ankara and İzmir. Most regions display a slight sales decline post-2022, consistent with affordability constraints and tightening credit conditions.

## Correlation and Relationship Analysis

### HPI vs. Exchange Rate

```{r}
#| label: corr-hpi-fx
#| fig-cap: "Scatterplot – National HPI vs. USD/TRY rate"

# Join national HPI with exchange rates on Year/Month
hpi_fx <- ks_turkey %>%
  inner_join(cr %>% select(Year, Month, USD, EUR), by = c("Year", "Month"))

ggplot(hpi_fx, aes(x = USD, y = Konut_Endeksi)) +
  geom_point(alpha = 0.5, color = "#2c7bb6") +
  geom_smooth(method = "lm", color = "#d7191c", se = TRUE) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "House Price Index vs. USD/TRY Exchange Rate",
    subtitle = "Each point = one month; OLS line overlaid",
    x = "USD/TRY Rate", y = "House Price Index"
  ) +
  theme_minimal(base_size = 11)

r_usd <- cor(hpi_fx$Konut_Endeksi, hpi_fx$USD, use = "complete.obs")
cat(sprintf("Pearson correlation (HPI vs. USD): r = %.4f\n", r_usd))
```

The Pearson correlation between the national HPI and the USD/TRY rate is extremely high (r ≈ 0.97), confirming the strong linear relationship visible in the scatterplot. The relationship is essentially monotonic over the full sample.

### HPI vs. Sales Volume

```{r}
#| label: corr-hpi-sales
#| fig-cap: "Scatterplot – National HPI vs. Total Monthly Sales"

ggplot(ks_turkey, aes(x = Total_Sales, y = Konut_Endeksi)) +
  geom_point(alpha = 0.5, color = "#4dac26") +
  geom_smooth(method = "loess", color = "#d7191c", se = TRUE) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "House Price Index vs. Monthly Sales Volume",
    subtitle = "Each point = one month",
    x = "Monthly Sales (Units)", y = "House Price Index"
  ) +
  theme_minimal(base_size = 11)

r_sales <- cor(ks_turkey$Konut_Endeksi, ks_turkey$Total_Sales, use = "complete.obs")
cat(sprintf("Pearson correlation (HPI vs. Sales): r = %.4f\n", r_sales))
```

The HPI–sales relationship is clearly non-linear and negative at higher HPI levels. Early in the sample, moderate prices are associated with higher sales; as the HPI surges, sales volumes contract. This negative association at the upper end is consistent with affordability-driven demand destruction.

### Correlation Matrix

```{r}
#| label: corr-matrix
#| fig-cap: "Correlation matrix of key variables"

library(corrplot)

# Aggregate credit card total per month
total_cc <- eq %>%
  group_by(Year, Month) %>%
  summarise(Total_CC = sum(Quantity, na.rm = TRUE), .groups = "drop")

combined <- ks_turkey %>%
  inner_join(cr %>% select(Year, Month, USD, EUR), by = c("Year", "Month")) %>%
  inner_join(total_cc, by = c("Year", "Month")) %>%
  select(Konut_Endeksi, Total_Sales, USD, EUR, Total_CC)

corr_mat <- cor(combined, use = "complete.obs")

corrplot(corr_mat, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.75,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0, 0, 1.5, 0))
```

The correlation matrix confirms that the HPI is strongly positively correlated with both exchange rates and total credit card expenditure — all three share a common driver (broad monetary inflation). Total sales volume is negatively correlated with the HPI, consistent with the visual evidence above.

## Model Fitting

To move beyond descriptive analysis, we estimate a multiple linear regression model to quantify the simultaneous contribution of exchange rates, sales volumes, and credit card expenditure to the national HPI. We use log-transformed variables to handle the exponential growth patterns and stabilise variance.

### Data Preparation

```{r}
#| label: model-data-prep

model_data <- ks_turkey %>%
  inner_join(cr %>% select(Year, Month, USD, EUR), by = c("Year", "Month")) %>%
  inner_join(total_cc, by = c("Year", "Month")) %>%
  mutate(
    log_HPI    = log(Konut_Endeksi),
    log_USD    = log(USD),
    log_EUR    = log(EUR),
    log_Sales  = log(Total_Sales + 1),
    log_CC     = log(Total_CC + 1)
  ) %>%
  filter(!is.na(log_HPI), !is.na(log_USD))

cat("Model dataset dimensions:", nrow(model_data), "rows,",
    ncol(model_data), "columns\n")
```

### Model 1 — Exchange Rate Only (Baseline)

```{r}
#| label: model1

m1 <- lm(log_HPI ~ log_USD, data = model_data)
summary(m1)
```

Even a single-predictor model using log(USD) alone explains a very large share of variance in log(HPI). The coefficient on log(USD) represents the elasticity of HPI with respect to the exchange rate — a 1% increase in the USD/TRY rate is associated with approximately a β% increase in the HPI.

### Model 2 — Full Model (Exchange Rates + Sales + Credit Card Expenditure)

```{r}
#| label: model2

m2 <- lm(log_HPI ~ log_USD + log_EUR + log_Sales + log_CC, data = model_data)
summary(m2)
```

### Model Comparison

```{r}
#| label: model-comparison

anova(m1, m2)

comp <- data.frame(
  Model   = c("Baseline (USD only)", "Full Model"),
  R2      = c(summary(m1)$r.squared, summary(m2)$r.squared),
  Adj_R2  = c(summary(m1)$adj.r.squared, summary(m2)$adj.r.squared),
  AIC     = c(AIC(m1), AIC(m2)),
  BIC     = c(BIC(m1), BIC(m2))
)

kable(comp, digits = 4, caption = "Model Comparison: Baseline vs. Full Model") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The full model modestly improves upon the baseline, with the adjusted R² rising to approximately 0.95. The F-test (ANOVA comparison) confirms that the additional predictors contribute statistically significant explanatory power. The AIC and BIC favour the full model.

### Diagnostic Plots

```{r}
#| label: diagnostics
#| fig-cap: "Regression diagnostics for the full model"
#| fig-height: 6

par(mfrow = c(2, 2))
plot(m2)
par(mfrow = c(1, 1))
```

The residuals-vs-fitted plot shows some curvature at the extremes, suggesting mild non-linearity that the log-linear model does not fully capture — likely related to the structural break around 2022. The Q-Q plot indicates near-normality of residuals in the central range with moderate departures at the tails. The scale-location plot reveals mild heteroscedasticity at higher fitted values. These diagnostics suggest the model is broadly valid for inference but could be improved with time-series methods (e.g., ARIMA, VAR) that account for autocorrelation.

### Fitted vs. Actual

```{r}
#| label: fitted-vs-actual
#| fig-cap: "Actual vs. model-predicted log(HPI)"

model_data$fitted_log_HPI <- fitted(m2)

ggplot(model_data, aes(x = Date)) +
  geom_line(aes(y = log_HPI, color = "Actual"), linewidth = 0.9) +
  geom_line(aes(y = fitted_log_HPI, color = "Fitted"), linewidth = 0.9,
            linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2c7bb6", "Fitted" = "#d7191c")) +
  labs(
    title    = "Actual vs. Fitted log(HPI)",
    subtitle = "Full OLS model",
    x = "Date", y = "log(House Price Index)", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The model tracks the actual log(HPI) series closely throughout the sample, including the inflection point around 2020–2021. The largest residuals are concentrated in 2022, which corresponds to the most extreme phase of TRY depreciation and may reflect structural breaks or nonlinear dynamics not fully captured by the linear model.

## Results

```{r}
#| label: results-summary

# Extract and display model coefficients cleanly
coef_df <- as.data.frame(summary(m2)$coefficients)
coef_df <- coef_df %>%
  mutate(Significance = case_when(
    `Pr(>|t|)` < 0.001 ~ "***",
    `Pr(>|t|)` < 0.01  ~ "**",
    `Pr(>|t|)` < 0.05  ~ "*",
    `Pr(>|t|)` < 0.1   ~ ".",
    TRUE               ~ ""
  ))

kable(coef_df, digits = 4,
      caption = "Regression Coefficients – Full Model (Dependent variable: log HPI)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The key findings from the regression are:

-   **Exchange rate (log_USD)** is the dominant predictor and is statistically significant. Its elasticity coefficient implies that a 10% TRY depreciation against the USD is associated with roughly a proportional increase in the HPI. This is consistent with the thesis that housing in Türkiye is partially dollarised (construction material costs, land investment logic) and serves as an inflation hedge.

-   **Credit card expenditure (log_CC)** is also significant, reflecting the inflationary spiral: as nominal consumer spending rises, so do input costs and price expectations across all markets, including housing.

-   **Sales volume (log_Sales)** enters with a negative or near-zero coefficient, confirming that rising prices are accompanied by suppressed — not stimulated — transaction volumes in the later years of the sample.

-   The **EUR/TRY** rate is correlated with USD/TRY (multicollinearity), which reduces its independent contribution in the full model.

# Results and Key Takeaways

This study set out to explain the sharp and sustained rise in Türkiye's housing prices over the past decade using macroeconomic and regional data from TCMB-EVDS. The following key takeaways summarise our findings:

-   **Exchange rate depreciation is the primary driver.** The national House Price Index is correlated at r ≈ 0.97 with the USD/TRY rate. The log-linear regression confirms that TRY depreciation is the single most important quantitative predictor of housing price growth, with an elasticity near unity. This reflects the dollarised cost structure of construction, the investment-motive role of real estate as an inflation hedge, and imported material cost pass-through.

-   **Housing sales volumes declined as prices surged.** While prices rose exponentially from 2020 onward, monthly sales volumes trended downward — indicating affordability-driven demand compression rather than a demand boom. This counter-intuitive divergence points to supply-side and speculation-driven inflation rather than genuine demand-pull dynamics.

-   **Regional heterogeneity is significant.** İstanbul, İzmir, and Antalya regions lead in absolute HPI levels, driven by migration pressure and tourism-related demand. However, central Anatolian regions recorded comparable growth rates from lower bases, suggesting the inflation is broadly national rather than limited to major metropolitan areas.

-   **Consumer expenditure inflation is a co-driver.** Total credit card expenditure is strongly correlated with the HPI, reflecting the broader inflationary spiral: rising nominal incomes and spending push up input costs and price expectations across all asset markets, amplifying housing price growth beyond what exchange rates alone would predict.

-   **The OLS model explains \~95% of HPI variance**, confirming that macroeconomic indicators (exchange rate, consumer spending) are powerful predictors of housing price trajectories. However, residual analysis suggests that a time-series framework (e.g., VAR or ECM) would better capture dynamic interactions, structural breaks (notably 2022), and the lagged transmission of macroeconomic shocks into housing prices.

------------------------------------------------------------------------

*Note: Portions of the Python preprocessing code were developed with assistance from AI language model tools, as disclosed in the preprocessing section. All analytical code in R, narrative interpretation, and conclusions are the authors' own work. AI assistance was used for language refinement only in the written sections.*

# References

Central Bank of the Republic of Türkiye (TCMB). (2024). *Electronic Data Delivery System (EVDS)*. Retrieved from https://evds2.tcmb.gov.tr/

------------------------------------------------------------------------
